Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-01 paper/zhang-cook-laa-langrene-menendez.tex
SA -> CRS, SAJO -> SA

% !TeX root = RJwrapper.tex
\title{Visual Diagnostics for Constrained Optimisation with Application to
Guided Tours}
\author{by H.Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas Langrené, Patricia Menéndez}

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: simulated annealing (SA), simulated annealing
with jump out (SAJO), and pseudo derivative (PD). Simulated annealing
(SA) is a random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search}.
\citet{posse1995projection} has proposed a slightly different cooling
scheme by introducing a halving parameter \(c\). In his proposal, the
\(\alpha\) is only adjusted if the last iteration takes more than \(c\)
times to find a better basis.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{input}{input}
  \SetKwInOut{output}{output}
    \input{$f(.)$, $\alpha_1$, $l_{\max}$, $\text{cooling}$} 
    \output{$\mathbf{A}_{l}$}
    Generate random start $\mathbf{A}_1$ and set $\mathbf{A}_{\text{cur}} \coloneqq \mathbf{A}_1$, $I_{\text{cur}} = f(\mathbf{A}_{\text{cur}})$, $j = 1$\;
  \Repeat{$\mathbf{A}_l$ is too close to $\mathbf{A}_{\text{cur}}$ in terms of geodesic distance}{
   Set $l = 1$\;
  \Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    Update $l = l + 1$\;
  }
  Update $\alpha_{j+1} = \alpha_j * \text{cooling}$\;
  Construct the geodesic interpolation between $\mathbf{A}_{\text{cur}}$ and $\mathbf{A}_l$\; 
  Update $\mathbf{A}_{\text{cur}} = \mathbf{A}_l$ and $j = j + 1$\;
}
  \caption{Simulated annealing (SA)}
  \label{random-search}
\end{algorithm}

Simulated annealing with jump out (SAJO)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as SA but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} highlights how SAJO differs from SA in the
inner loop.

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$ or $P > U$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$, $T(l) = \frac{T_0}{\log(l + 1)}$ and $P= \min\left\{\exp\left[-\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
    Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
    Update $l = l + 1$\;
  }
  \caption{Simulated annealing with jump out (SAJO)}
  \label{simulated-annealing}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit
optimisers: 1) projection bases \(\mathbf{A}\), 2) index values \(I\),
and 3) state \(S\). For SA and SAJO, possible states include
\texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. Pseudo derivative (PD) has a wider variety of
states including \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}. Multiple iterators index the information
collected at different levels: \(t\) is a unique identifier prescribing
the natural ordering of each observation; \(j\) and \(l\) are the
counter of the outer and inner loop respectively. Other parameters of
interest recorded include \texttt{method} that tags the name of the
optimiser, and \texttt{alpha} that indicates the sampling neighbourhood
size for searching observations. A matrix notation of the data structure
is presented in equation \ref{eq:data-structure}.

@@

Figure \ref{fig:toy-search} shows an example of the search plot for SA
(left) and SAJO (right). Both optimisers quickly find better bases in
the first few iterations and then take longer to find one in the later
iterations. The anchor bases, the ones found with the highest index
value in each iteration, always have an increased index value in the
optimiser SA while this is not the case for SAJO. This feature gives SA
an advantage in this simple example to quickly find the optimum.

@@

\caption[A comparison of the searches by two optimisers]{A comparison of the searches by two optimisers: SA (left) and SAJO (right) on a 2D projection problem of a six-variable dataset, \code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SAJO longer to find the final basis.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SAJO (right) is observed to first pass by some bases with higher index value before reach the target bases in Time 76-130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting diagnostic is to examine how the index value
improves between interpolating bases since the projection on these bases
is played by the tour animation. Trace plots are created by plotting the
index value against time. Figure \ref{fig:toy-interp} presents the trace
plot of the same optimisers as Figure \ref{fig:toy-search} and one can
observe that the trace is smooth in both cases. It may seem bizarre at
the first sight that the interpolation in the left panel first passes
some bases with higher index value before decreases to a lower target.
This happens because on the one hand, the probabilistic acceptance in
SAJO implies that some worse bases will be accepted by the optimiser. On
the other hand, guided tour interpolate between the current and target
basis to provide a smooth transition between projections. These two
facts indicate that a non-monotonic interpolation can't be avoided for
SAJO. Later, in section \ref{monotonic}, there will be a discussion on
improving the non-monotonic interpolation for SA.

@@

\caption[Search paths of SA (green) and PD (brow) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index]{Search paths of SA (green) and PD (brow) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish are better identified. SA (green) finishes the optimisation quicker than PD (brown) since there is no further movement for SA in the sixth frame. The full video of the animation can be found at \url{https://vimeo.com/504242845}.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

Animation is another type of display to show how the search progresses
from start to finish in the space. An \texttt{animate\ =\ TRUE} argument
is used to enable the animation in the PCA plot. Figure
\ref{fig:toy-pca-animated} shows six frames from the animation of the
PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information
one can learn from this animation is that SA finds its end basis quicker
than PD since SA finishes its search in the 5th frame while PD is still
making more progress.

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, the space
of \(p \times 2\) bases is a \(p\)-D torus. If take \(p = 3\), one would
see a classical 3D torus shape as shown by the grey points in Figure
\ref{fig:toy-torus}. The two circles of the torus can be observed to be
perpendicular to each other and this can be linked back to the
orthogonality of the columns in the bases. Two paths optimised by SA and
PD are plotted on top of the torus and coloured in green and brown
respectively to match the previous few 1D space plots. The final basis
found by PD and SA are shown in a larger shape and printed below
respectively:

@@

\caption{Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and SA (green). The projection problem is a 2D projection with three variables using holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \url{https://vimeo.com/512882784}.}\label{fig:toy-torus}
\end{figure}
\end{Schunk}

@@

An example of non-monotonic interpolation has been given in Figure
\ref{fig:toy-interp}: a path that passes bases with higher index value
than the target one. For SAJO, a non-monotonic interpolation is
justified since target bases do not necessarily have a higher index
value than the current one, while this is not the case for SA. The
original trace plot for a 2D projection problem, optimised by SA, is
shown on the left panel of Figure \ref{fig:interruption} and one can
observe clearly that the non-monotonic interpolation has undermined the
optimiser to realise its full potential. Hence, an interruption is
implemented to stop at the best basis found in the interpolation. The
right panel of Figure \ref{fig:interruption} shows the trace plot after
implementing the interruption and while the first two interpolations are
identical, the basis at Time 61 has a higher index value than the target
in the third interpolation. Rather than starting the next iteration from
the target basis on Time 65, SA starts the next iteration at Time 61 on
the right panel and reaches a better final basis.

@@

\caption[Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by SA]{Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by SA. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

\caption[Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \code{boa6} data using holes index]{Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \code{boa6} data using holes index. Optimiser SA is used before the polish. The clustering structure in the data is detected by SA but the polish step improves the index value and produces clearer boundaries of the clusters, espeically along the vertial axis.}\label{fig:polish}
\end{figure}
\end{Schunk}

Once the final basis has been found by an optimiser, one may want to
push further in the close neighbourhood to see if an even better basis
can be found. A polish search takes the final basis of an optimiser as
the start of a new guided tour to search for local breakthrough. The
polish algorithm is similar to the SA but with three distinctions: 1) a
hundred rather than one candidate bases are generated each time in the
inner loop; 2) the neighbourhood size is reduced in the inner loop,
rather than in the outer loop; and 3) three more termination conditions
have been added to ensure the new basis generated is distinguishable
from the current one in terms of the distance in the space, percentage
change in the index value, and neighbourhood size:

@@

\caption[Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \code{boa5}]{Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \code{boa5}. Both SA and SAJO succeeds in the optimisation, PD fails to optimise this non-smooth index. Further, SAJO takes  much longer than SA to finish the optimisation, it finishes off closer to the theoretical best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of
all three optimisers and as expected, none of the interpolated path is
smooth. There is barely any improvement made by PD, indicating its
failure in optimising non-smooth index. While SA and SAJO have managed
to get close to the index value of the theoretical best, trace plot
shows that it takes SAJO much longer to interpolate towards the final
basis. This long interpolation path is partially due to the fluctuation
in the early iterations, where the SAJO tends to generously accept
inferior bases before concentrating near the optimum. The PCA plot shows
the interpolation path and search points excluding the last termination
iteration. Pseudo Derivative (PD) quickly gets stuck near the starting
position. Comparing the amount of random search done by SA and SAJO, it
is surprising that SAJO does not carry as many samples as SA. Combining
the insights from both the trace and PCA plot, one can learn the two
different search strategies by SA and SAJO: SA frequently samples in the
space and only make a move when an improvement is guaranteed to be made,
while SAJO first broadly accepts bases in the space and then starts the
extensive sampling in a narrowed subspace. The specific decision of
which optimiser to use will depend on the index curve in the basis space
but if the basis space is non-smooth, accepting inferior bases at first,
as what SAJO has done, can lead to a more efficient search, in terms of
the overall number of points evaluated.

@@

\caption[Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers]{Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: SA and SAJO, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projecton index, $I^{nk}(n)$, for \code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*).}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares the performance of SA and SAJO when a local
maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared
to understand how a large search neighbourhood would affect the final
basis and the length of the search. Figure \ref{fig:kol-result} shows 80
paths simulated using 20 seeds in the PCA plot, faceted by optimiser and
search size. With SA and a search size of 0.5, despite being the
simplest and fastest, the optimiser fails in three instances where the
final basis lands neither near the local nor the global optimum. With a
larger search size of 0.7, more seeds have found the global maximum.
Comparing SA and SAJO for a search size of 0.5, SAJO does not seem to
improve the final basis found, despite having longer interpolation
paths. However, the denser paths near the local maximum is an indicator
that SAJO is working hard to examine if there is any other optimum in
the basis space but the relative small search size has diminished its
ability to reach the global maximum. With a larger search size, almost
all the seeds (16 out of 20) have found the global maximum and some
final bases are much closer to the theoretical best, as compared to the
three other cases. This indicates that SAJO, with a reasonable large
search size, is able to overcome the local optimum and optimise close
towards the global optimum.

@@

\caption[Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis]{Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis. Optimisation is on the 1D projection index, $I^{nk}(n)$, for boa6 data using SA with seed 2463. The dots represent the target basis in each iteration and the path shows the interpolation. On the left panel, one target basis is generated with an opposite orientation to the current basis (hence appear on the other side of the basis space) and the interpolator crosses the origin to perform the interpolation. The right panel shows the same interpolation after implementing an orientation check and the undesirable interpolation disappears.}\label{fig:flip-sign}
\end{figure}
\end{Schunk}

@@

\address{%
H.Sherry Zhang\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}

\address{%
Dianne Cook\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}

\address{%
Ursula Laa\\
University of Natural Resources and Life Sciences\\
Institute of Statistics\\
}
\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}

\address{%
Nicolas Langrené\\
CSIRO Data61\\
34 Village Street, Docklands VIC 3008 Australia\\
}
\href{mailto:nicolas.langrene@csiro.au}{\nolinkurl{nicolas.langrene@csiro.au}}

\address{%
Patricia Menéndez\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}

>>>>

% !TeX root = RJwrapper.tex
\title{Visual Diagnostics for Constrained Optimisation with Application to
Guided Tours}
\author{by H.Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas Langrené, and Patricia Menéndez}

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: Creeping random search (CRS), simulated annealing
(SA), and pseudo derivative (PD). Creeping random search (CRS) is a
random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search}.
\citet{posse1995projection} has proposed a slightly different cooling
scheme by introducing a halving parameter \(c\). In his proposal, the
\(\alpha\) is only adjusted if the last iteration takes more than \(c\)
times to find a better basis.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{input}{input}
  \SetKwInOut{output}{output}
    \input{$f(.)$, $\alpha_1$, $l_{\max}$, $\text{cooling}$} 
    \output{$\mathbf{A}_{l}$}
    Generate random start $\mathbf{A}_1$ and set $\mathbf{A}_{\text{cur}} \coloneqq \mathbf{A}_1$, $I_{\text{cur}} = f(\mathbf{A}_{\text{cur}})$, $j = 1$\;
  \Repeat{$\mathbf{A}_l$ is too close to $\mathbf{A}_{\text{cur}}$ in terms of geodesic distance}{
   Set $l = 1$\;
  \Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    Update $l = l + 1$\;
  }
  Update $\alpha_{j+1} = \alpha_j * \text{cooling}$\;
  Construct the geodesic interpolation between $\mathbf{A}_{\text{cur}}$ and $\mathbf{A}_l$\; 
  Update $\mathbf{A}_{\text{cur}} = \mathbf{A}_l$ and $j = j + 1$\;
}
  \caption{Creeping random search (CRS)}
  \label{random-search}
\end{algorithm}

Simulated annealing (SA)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as CRS but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} highlights how SAJO differs from SA in the
inner loop.

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$ or $P > U$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$, $T(l) = \frac{T_0}{\log(l + 1)}$ and $P= \min\left\{\exp\left[-\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
    Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
    Update $l = l + 1$\;
  }
  \caption{Simulated annealing (SA)}
  \label{simulated-annealing}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit
optimisers: 1) projection bases \(\mathbf{A}\), 2) index values \(I\),
and 3) state \(S\). For CRS and SA, possible states include
\texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. Pseudo derivative (PD) has a wider variety of
states including \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}. Multiple iterators index the information
collected at different levels: \(t\) is a unique identifier prescribing
the natural ordering of each observation; \(j\) and \(l\) are the
counter of the outer and inner loop respectively. Other parameters of
interest recorded include \texttt{method} that tags the name of the
optimiser, and \texttt{alpha} that indicates the sampling neighbourhood
size for searching observations. A matrix notation of the data structure
is presented in equation \ref{eq:data-structure}.

@@

Figure \ref{fig:toy-search} shows an example of the search plot for CRS
(left) and SA (right). Both optimisers quickly find better bases in the
first few iterations and then take longer to find one in the later
iterations. The anchor bases, the ones found with the highest index
value in each iteration, always have an increased index value in the
optimiser CRS while this is not the case for SA. This feature gives SA
an advantage in this simple example to quickly find the optimum.

@@

\caption[A comparison of the searches by two optimisers]{A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\caption[An inspection of the index values as the optimisation progress for two optimisers]{An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reach the target bases in Time 76-130.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting diagnostic is to examine how the index value
improves between interpolating bases since the projection on these bases
is played by the tour animation. Trace plots are created by plotting the
index value against time. Figure \ref{fig:toy-interp} presents the trace
plot of the same optimisers as Figure \ref{fig:toy-search} and one can
observe that the trace is smooth in both cases. It may seem bizarre at
the first sight that the interpolation in the left panel first passes
some bases with higher index value before decreases to a lower target.
This happens because on the one hand, the probabilistic acceptance in SA
implies that some worse bases will be accepted by the optimiser. On the
other hand, guided tour interpolate between the current and target basis
to provide a smooth transition between projections. These two facts
indicate that a non-monotonic interpolation can't be avoided for SA.
Later, in section \ref{monotonic}, there will be a discussion on
improving the non-monotonic interpolation for CRS.

@@

\caption[Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index]{Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish are better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \url{https://vimeo.com/504242845}.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

Animation is another type of display to show how the search progresses
from start to finish in the space. An \texttt{animate\ =\ TRUE} argument
is used to enable the animation in the PCA plot. Figure
\ref{fig:toy-pca-animated} shows six frames from the animation of the
PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information
one can learn from this animation is that CRS finds its end basis
quicker than PD since CRS finishes its search in the 5th frame while PD
is still making more progress.

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, the space
of \(p \times 2\) bases is a \(p\)-D torus. If take \(p = 3\), one would
see a classical 3D torus shape as shown by the grey points in Figure
\ref{fig:toy-torus}. The two circles of the torus can be observed to be
perpendicular to each other and this can be linked back to the
orthogonality of the columns in the bases. Two paths optimised by CRS
and PD are plotted on top of the torus and coloured in green and brown
respectively to match the previous few 1D space plots. The final basis
found by PD and CRS are shown in a larger shape and printed below
respectively:

@@

\caption{Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \url{https://vimeo.com/512882784}.}\label{fig:toy-torus}
\end{figure}
\end{Schunk}

@@

An example of non-monotonic interpolation has been given in Figure
\ref{fig:toy-interp}: a path that passes bases with higher index value
than the target one. For SA, a non-monotonic interpolation is justified
since target bases do not necessarily have a higher index value than the
current one, while this is not the case for CRS. The original trace plot
for a 2D projection problem, optimised by CRS, is shown on the left
panel of Figure \ref{fig:interruption} and one can observe clearly that
the non-monotonic interpolation has undermined the optimiser to realise
its full potential. Hence, an interruption is implemented to stop at the
best basis found in the interpolation. The right panel of Figure
\ref{fig:interruption} shows the trace plot after implementing the
interruption and while the first two interpolations are identical, the
basis at Time 61 has a higher index value than the target in the third
interpolation. Rather than starting the next iteration from the target
basis on Time 65, CRS starts the next iteration at Time 61 on the right
panel and reaches a better final basis.

@@

\caption[Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS]{Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \code{boa6} data using holes index, optimised by CRS. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel.}\label{fig:interruption}
\end{figure}
\end{Schunk}

@@

\caption[Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \code{boa6} data using holes index]{Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \code{boa6} data using holes index. Optimiser CRS is used before the polish. The clustering structure in the data is detected by CRS but the polish step improves the index value and produces clearer boundaries of the clusters, espeically along the vertial axis.}\label{fig:polish}
\end{figure}
\end{Schunk}

Once the final basis has been found by an optimiser, one may want to
push further in the close neighbourhood to see if an even better basis
can be found. A polish search takes the final basis of an optimiser as
the start of a new guided tour to search for local breakthrough. The
polish algorithm is similar to the CRS but with three distinctions: 1) a
hundred rather than one candidate bases are generated each time in the
inner loop; 2) the neighbourhood size is reduced in the inner loop,
rather than in the outer loop; and 3) three more termination conditions
have been added to ensure the new basis generated is distinguishable
from the current one in terms of the distance in the space, percentage
change in the index value, and neighbourhood size:

@@

\caption[Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \code{boa5}]{Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \code{boa5}. Both CRS and SA succeeds in the optimisation, PD fails to optimise this non-smooth index. Further, SA takes  much longer than CRS to finish the optimisation, it finishes off closer to the theoretical best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of
all three optimisers and as expected, none of the interpolated path is
smooth. There is barely any improvement made by PD, indicating its
failure in optimising non-smooth index. While CRS and SA have managed to
get close to the index value of the theoretical best, trace plot shows
that it takes SA much longer to interpolate towards the final basis.
This long interpolation path is partially due to the fluctuation in the
early iterations, where the SA tends to generously accept inferior bases
before concentrating near the optimum. The PCA plot shows the
interpolation path and search points excluding the last termination
iteration. Pseudo Derivative (PD) quickly gets stuck near the starting
position. Comparing the amount of random search done by CRS and SA, it
is surprising that SA does not carry as many samples as CRS. Combining
the insights from both the trace and PCA plot, one can learn the two
different search strategies by CRS and SA: CRS frequently samples in the
space and only make a move when an improvement is guaranteed to be made,
while SA first broadly accepts bases in the space and then starts the
extensive sampling in a narrowed subspace. The specific decision of
which optimiser to use will depend on the index curve in the basis space
but if the basis space is non-smooth, accepting inferior bases at first,
as what SA has done, can lead to a more efficient search, in terms of
the overall number of points evaluated.

@@

\caption[Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers]{Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: CRS and SA, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projecton index, $I^{nk}(n)$, for \code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*).}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares the performance of CRS and SA when a local
maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared
to understand how a large search neighbourhood would affect the final
basis and the length of the search. Figure \ref{fig:kol-result} shows 80
paths simulated using 20 seeds in the PCA plot, faceted by optimiser and
search size. With CRS and a search size of 0.5, despite being the
simplest and fastest, the optimiser fails in three instances where the
final basis lands neither near the local nor the global optimum. With a
larger search size of 0.7, more seeds have found the global maximum.
Comparing CRS and SA for a search size of 0.5, SA does not seem to
improve the final basis found, despite having longer interpolation
paths. However, the denser paths near the local maximum is an indicator
that SA is working hard to examine if there is any other optimum in the
basis space but the relative small search size has diminished its
ability to reach the global maximum. With a larger search size, almost
all the seeds (16 out of 20) have found the global maximum and some
final bases are much closer to the theoretical best, as compared to the
three other cases. This indicates that SA, with a reasonable large
search size, is able to overcome the local optimum and optimise close
towards the global optimum.

@@

\caption[Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis]{Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis. Optimisation is on the 1D projection index, $I^{nk}(n)$, for boa6 data using CRS with seed 2463. The dots represent the target basis in each iteration and the path shows the interpolation. On the left panel, one target basis is generated with an opposite orientation to the current basis (hence appear on the other side of the basis space) and the interpolator crosses the origin to perform the interpolation. The right panel shows the same interpolation after implementing an orientation check and the undesirable interpolation disappears.}\label{fig:flip-sign}
\end{figure}
\end{Schunk}

@@

\address{%
H.Sherry Zhang\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\\\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\\\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
}

\address{%
Ursula Laa\\
University of Natural Resources and Life Sciences\\%
Institute of Statistics\\
%
%
%
\\\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}
}

\address{%
Nicolas Langrené\\
CSIRO Data61\\%
34 Village Street, Docklands VIC 3008 Australia\\
%
%
%
\\\href{mailto:nicolas.langrene@csiro.au}{\nolinkurl{nicolas.langrene@csiro.au}}
}

\address{%
Patricia Menéndez\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\\\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}
}
